# Workflow to process species occurrences

## Set Up

```{r}
source("setup.R")
```

## Explore and Clean Survey Data

Read in raw tables from Access database (PC version)

```{r}
# NOTE: THIS ONLY WORKS AS IS ON A PC, will likely throw errors on a Mac
# data stored as an Access database
library(RODBC)

# read in Access object
survey_db <- odbcConnectAccess2007("data/nature_in_the_city/nic_bird_butterfly_surveys_2018.mdb")

# get a list of all tables
sqlTables(survey_db)


# read in necessary files
bird_obs <- sqlFetch(survey_db, "bird_observations") %>% as_tibble()

butterfly_obs <- sqlFetch(survey_db, "butterfly_observations")

sites <- sqlFetch(survey_db, "sites")


```

Map out sites

```{r}
sites %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  mapview()
```

Filter species of interest

*For now, focus on same species used in 2019 Nature in the City Report:*

+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| **Species**                 | **Relation to ecological metrics and possible ecological indications**                                                      |
+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Red-winged Blackbird (RWBL) | Negatively correlated with detections of urban-adapted bird species.                                                        |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             | The presence of this species may indicate that the community and habitat are not suitable for urban-adapted bird species.   |
+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Western Kingbird (WEKI)     | Positively correlated with detections urban-sensitive bird species.                                                         |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             | The presence of this species may indicate that the community and habitat can support urban-sensitive bird species.          |
+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Western Meadowlark (WEME)   | Positively correlated with detections urban-sensitive bird species.                                                         |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             | The presence of this species may indicate that the community and habitat can support urban-sensitive bird species.          |
+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Yellow Warbler (YEWA)       | Species of interest for the City of Fort Collins Natural Areas program.                                                     |
+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Clouded Sulphur             | Positively correlated with detections of non-resident butterfly species.                                                    |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             | The presence of this species may indicate that the community and habitat can support a non-resident butterfly species.      |
+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Orange Sulphur              | Positively correlated with butterfly species richness.                                                                      |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             | The presence of this species may indicate that the community and habitat can support a wide variety of butterfly species.   |
+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Painted Lady                | Positively correlated with butterfly species richness.                                                                      |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             |                                                                                                                             |
|                             | The presence of this species may indicate that the community and habitat can support a wide variety of butterfly species.   |
+-----------------------------+-----------------------------------------------------------------------------------------------------------------------------+

```{r}
# double-check all species are spelled correctly
butterfly_obs %>% distinct(species) %>% View()

bird_obs %>% distinct(species) %>% View()
# WEME has two distinct obs, likely an extra species. Remove spacies in all species name entires below


obs_final <- bird_obs %>% 
  # remove spaces in bird names
  mutate(species = str_replace(species, " ", "")) %>% 
  # combine bird and butterfly datasets
  bind_rows(
    butterfly_obs) %>% 
  filter(species %in% c("RWBL", "WEKI", "WEME", "YEWA", "Clouded Sulfur", "Orange Sulfur", "Painted Lady")) %>% 
  # only keep subset of variables
  select(observation_ID, survey_ID, site_ID, site_visit, species, habitat_type) %>%   # combine with site data
  left_join(sites, by = "site_ID")

# save this file for future use (remove need to use Access data table)
write_csv(obs_final, "data/NIC_survey_occ_processed.csv")
```

## Read and Process Additional Occurrence Data
