---
title: "Code to process predictors for urban niche model"
author: "Mikko Jimenez, Josh Carrell, and Caitlin Mothes"
date: "'r Sys.Date()'"
output: html_notebook
---

## Process predictors

As part of a broader pipeline for running a niche model in urban settings, this script is designed to provide a flexible process for reading in predictor variables, setting them to a consistent coordinate reference system, and cropping them to an extent of interest. 

Note: it is currently unclear if predictors will be expected to be downloaded prior to running this script or if we will be pulling predictors directly from an API. I've written code for each situation below. 

## set up environment
```{r}
source("setup.R") 

# read in spatial layer that includes full extent of interest, in this case a boundary .shp
template_path <- "data/nature_in_the_city/gis/ft_collins_GMA_boundary.shp"

# list of paths to predictors (shapefiles, .tifs, etc.)
predictor_paths <- c("data/input_raw/Hydrology/Hydrology.shp",
                     "data/input_raw/nlcd_tcc_CONUS_2021_v2021-4/nlcd_tcc_conus_2021_v2021-4.tif")

# set desired resolution
resolution = 100

# set output directory
output_path <- "data/input_processed"
```

## set the 'template' crs and extent and write a function that matches these for predictors
```{r}
# write function that reads in predictors and matches CRS and extent to the template
process_file <- function(input_path, template_path, resolution, save = FALSE, output_path = "data/input_processed") {
  # Read in the template .shp and save as an sf object
  template <- read_sf(template_path)
  assign("template_sf", template, envir = .GlobalEnv)
  
  # Initialize variables to hold the processed objects
  rast_shapefile <- NULL
  raster_file <- NULL

  if (grepl("\\.shp$", input_path, ignore.case = TRUE)) {
    # Read in shapefile
    shapefile <- st_read(input_path)

    # Set CRS to match the template
    shapefile <- st_transform(shapefile, crs = crs(template))

    # Set extent to match the template (crop if necessary)
    shapefile <- st_crop(shapefile, st_bbox(template))
    
    ## Rasterize the shapefile
    # Create an empty raster template with the specified resolution
    raster_template <- rast(ext(template), resolution = resolution)
    # Rasterize shapefile 
    rast_shapefile <- terra::rasterize(shapefile, raster_template, fun = mean)
    # Set CRS to match the shapefile
    crs(rast_shapefile) <- st_crs(shapefile)$proj4string

    # Extract filename without extension for naming
    file_name <- tools::file_path_sans_ext(basename(input_path))

    # Save as an independent object
    assign(file_name, rast_shapefile, envir = .GlobalEnv)
    
  } else if (grepl("\\.tif$", input_path, ignore.case = TRUE)) {
    # Read in raster and template files
    raster_file <- rast(input_path)

    # Set template temporarily to raster CRS
    shapefile_temp <- st_transform(template, crs = crs(raster_file))
    raster_file <- crop(raster_file, shapefile_temp)
    
    # Project cropped raster into template CRS
    raster_file <- project(raster_file, crs(template), res = resolution)
    
    # Extract filename without extension for naming
    file_name <- tools::file_path_sans_ext(basename(input_path))
    
    # Save as an independent object
    assign(file_name, raster_file, envir = .GlobalEnv)
    
  } else {
    stop("Unsupported file type.")
  }
  
  # Save the processed objects to the specified directory if 'save' is TRUE
  if (save) {
    if (!dir.exists(output_path)) {
      dir.create(output_path, recursive = TRUE)
    }
    
    if (!is.null(rast_shapefile)) {
      # Write raster file
      writeRaster(rast_shapefile, filename = file.path(output_path, paste0(file_name, ".tif")), overwrite = TRUE)
    }
    
    if (!is.null(raster_file)) {
      # Write raster file
      writeRaster(raster_file, filename = file.path(output_path, paste0(file_name, ".tif")), overwrite = TRUE)
    
    }
  }
}

# Process each predictor file and save as independent objects
for (input_path in predictor_paths) {
  process_file(input_path, template_path, resolution, save=TRUE, output_path)
}
```

## plot all layers in tmap to explore overlays 
```{r}
# plot the layers using tmap
tmap_mode("view")

# add to this as needed
tm_shape(template_sf) +
  tm_lines(col = "red", lwd = 2) +
  tm_shape(`nlcd_tcc_conus_2021_v2021-4`) +
  tm_raster(palette = "Greens", alpha = 0.5, title = "NLCD TCC 2021") +
  tm_shape(Hydrology) +
  tm_raster(palette = "Blues", alpha = 0.5, title = "Hydrology")
```


