# Workflow to process species occurrences

## Set Up

```{r}
source("setup.R")
```

## Explore and Clean Survey Data

Read in raw tables from Access database (PC version)

```{r}
# NOTE: THIS ONLY WORKS AS IS ON A PC, will likely throw errors on a Mac
# data stored as an Access database
library(RODBC)

# read in Access object
survey_db <- odbcConnectAccess2007("data/nature_in_the_city/nic_bird_butterfly_surveys_2018.mdb")

# get a list of all tables
sqlTables(survey_db)


# read in necessary files

## birds
bird_obs <- sqlFetch(survey_db, "bird_observations")

bird_species <- sqlFetch(survey_db, "bird_species")

## combine
birds <- left_join(bird_obs, bird_species, by = c("species" = "alpha_code"))

## butterflies
butterfly_obs <- sqlFetch(survey_db, "butterfly_observations")

butterfly_species <- sqlFetch(survey_db, "butterfly_species")

butterflies <- left_join(butterfly_obs, butterfly_species, by = c("species" = "common_name"))

## site info
sites <- sqlFetch(survey_db, "sites")




```

Map out sites

```{r}
sites %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  mapview()
```

Filter species of interest

*For now, focus on same species used in 2019 Nature in the City Report:*

+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+
| **Species**                 | **Relation to ecological metrics and possible ecological indications**                                                    |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+
| Red-winged Blackbird (RWBL) | Negatively correlated with detections of urban-adapted bird species.                                                      |
|                             |                                                                                                                           |
|                             | The presence of this species may indicate that the community and habitat are not suitable for urban-adapted bird species. |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+
| Western Kingbird (WEKI)     | Positively correlated with detections urban-sensitive bird species.                                                       |
|                             |                                                                                                                           |
|                             | The presence of this species may indicate that the community and habitat can support urban-sensitive bird species.        |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+
| Western Meadowlark (WEME)   | Positively correlated with detections urban-sensitive bird species.                                                       |
|                             |                                                                                                                           |
|                             | The presence of this species may indicate that the community and habitat can support urban-sensitive bird species.        |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+
| Yellow Warbler (YEWA)       | Species of interest for the City of Fort Collins Natural Areas program.                                                   |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+
| Clouded Sulphur             | Positively correlated with detections of non-resident butterfly species.                                                  |
|                             |                                                                                                                           |
|                             | The presence of this species may indicate that the community and habitat can support a non-resident butterfly species.    |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+
| Orange Sulphur              | Positively correlated with butterfly species richness.                                                                    |
|                             |                                                                                                                           |
|                             | The presence of this species may indicate that the community and habitat can support a wide variety of butterfly species. |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+
| Painted Lady                | Positively correlated with butterfly species richness.                                                                    |
|                             |                                                                                                                           |
|                             | The presence of this species may indicate that the community and habitat can support a wide variety of butterfly species. |
+-----------------------------+---------------------------------------------------------------------------------------------------------------------------+

```{r}
# double-check all species are spelled correctly
butterflies %>% distinct(scientific_name) %>% View()

birds %>% distinct(species) %>% View()
# WEME has two distinct obs, likely an extra species. Remove spaces in all species name entires below


obs_final <- birds %>% 
  # remove spaces in bird names
  mutate(species = str_replace(species, " ", "")) %>% 
  # combine bird and butterfly datasets
  bind_rows(
    butterflies) %>% 
  filter(species %in% c("RWBL", "WEKI", "WEME", "YEWA", "Clouded Sulphur", "Orange Sulphur", "Painted Lady")) %>% 
  # only keep subset of variables
  select(observation_ID, survey_ID, site_ID, site_visit, species, habitat_type, scientific_name) %>%   # combine with site data
  left_join(sites, by = "site_ID") %>% 
  #fix the one missing scientific name
  mutate(scientific_name = if_else(species == "WEME", "Sturnella neglecta", scientific_name)) %>% 
  # keep distinct records/site
  distinct(species, latitude, longitude, .keep_all = TRUE)

# save this file for future use (remove need to use Access data table)
write_csv(obs_final, "data/NIC_survey_occ_processed.csv")
```

## Read and Process Additional Occurrence Data

Read in aoi (FOCO GMA)

```{r}
foco_boundary <- read_sf("data/nature_in_the_city/gis/ft_collins_GMA_boundary.shp") %>%
  # convert boundary line to single polygon
  st_cast("POLYGON") %>%
  st_make_valid()


```

Get list of species names

```{r}
species_names <- read_csv("data/NIC_survey_occ_processed.csv") %>% 
  pull(scientific_name) %>% unique()
```

Pull and clean observations from `spocc`

```{r}
library(spocc)

# set up parameters

## get bounds for socc search
bounds <- foco_boundary %>% 
  # make sure in WGS 84
  st_transform(crs = 4326) %>% 
  # get bounds 
  st_bbox() %>% 
  as.vector()

# date max range (Sys.Date operates weird inside a vector)
max_date <- Sys.Date()

# can't get occ() filters to work so will do that after the original call
sp_occ <- spocc::occ(
  query = species_names[1],
  from = c(
    "gbif",
    "inat",
    "ecoengine",
    "vertnet",
    "bison",
    "ala",
    "idigbio",
    "ebird"
  ),
  geometry = bounds,
  #date = c("2014-01-01", max_date),
  has_coords = TRUE,
  limit = 5000
) %>%
  spocc::occ2df()

  
```

Compare total number of occurrences per source

```{r}
sp_occ %>% 
  group_by(prov) %>% 
  count()

# compare after removing duplicates
sp_occ %>% 
  group_by(prov) %>% 
  distinct(latitude, longitude) %>% count()
```

ebird needs an API key, maybe just stick with gbif for now based on results (definitely has the most occurrences)

Make spatial and map them out

```{r}
sp_occ_sf <- sp_occ %>% 
  # remove missing coords
  filter(!is.na(latitude), !is.na(longitude)) %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
```

Map

```{r}
sp_occ_sf %>% 
  st_filter(st_transform(foco_boundary, crs = st_crs(sp_occ_sf))) %>% 
tm_shape() +
  tm_dots()
```

See which provider did not filter - (GBIF and inat did, vertnet did not, as expected)

```{r}
sp_occ_sf %>% 
  st_filter(st_transform(foco_boundary, crs = st_crs(sp_occ_sf))) %>% 
  group_by(prov) %>% 
  count()
```
